---
title: Substation Analysis
output: html_document
author: "Craig Nicholson"
params:
  dayOfAnalysis: !r as.Date("2016-03-02")
  pfHigh: 0.99
  pfLow: 0.98
  voltageLow: 114.0
  voltageHigh: 126.0
  voltageLowLimit: 100.0
  voltageMaxLimit: 140.0
---

## Abstract
I will describe the votlage, megawatts, megavars (receieved and delivered),
and power factor (lagging and leading) in terms of the electricity. Key goals
are to understand substations and meters with regards to voltage
and power factor as well as megawatts and megavars.

## Introduction
Electric utilities collect meter readings in time intervals in various units. 
The intervals can be from 1 minute to 60 minute intervals collecing kW, kWh, 
VARh, volts.  An utility has to maintain a specified range for voltage across
their electric network.  When voltage is too low brown outs or electric motors
may fail to work and when voltage is too high appliances and equiment can 
overheat, burn up, and possibly explode.


### Power Factor
A negative power factor (Lagging) occurs when the device (which is normally 
the load) generates power, which then flows back towards the source, which 
is normally considered the generator.

In an electric power system, a load with a **low power** factor draws more 
current than a load with a high power factor for the same amount of useful 
power transferred. 

The higher currents increase the energy lost in the distribution system, 
and require larger wires and other equipment. Because of the costs of 
larger equipment and wasted energy, electrical utilities will usually 
charge a higher cost to industrial or commercial customers where there 
is a low power factor.

Power factors **below 1.0** require a utility to generate more than the 
minimum volt-amperes necessary to supply the real power (watts).  
This increases generation and transmission costs. For example, if the 
load power factor were as low as 0.7, the apparent power would be 1.4 
times the real power used by the load. Line current in the circuit would 
also be 1.4 times the current required at 1.0 power factor, so the losses 
in the circuit would be doubled 
(**since they are proportional to the square of the current**). 

Alternatively all components of the system such as generators, conductors, 
transformers, and switchgear would be increased in size (and cost) 
to carry the extra current.

Utilities typically charge additional costs to commercial customers who have 
a power factor below some limit, which is typically 0.9 to 0.95. Engineers 
are often interested in the power factor of a load as one of the factors that 
affect the efficiency of power transmission.

#### Definitions
Since the units are consistent, the power factor is by definition a dimensionless number between −1 and 1. When power factor is equal to 0, the energy flow is entirely reactive and stored energy in the load returns to the source on each cycle. When the power factor is 1, all the energy supplied by the source is consumed by the load.

Power factors are usually stated as "leading" or "lagging" to show the sign of the
phase angle. Capacitive loads are leading (current leads voltage) and supply power, and inductive loads are lagging (current lags voltage) and consume power.


```{r Setup knitr, echo=FALSE}
knitr::opts_chunk$set(include=TRUE, echo=FALSE, warning=FALSE, message=FALSE)
```

```{r Setup Libraries & packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(memisc)       # clashes with dplyr rename, and select
                      # and I am using limited funcationality
                      # loading first so memisc objects are masked
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(knitr)
library(lubridate)    # working with dates
```

```{r Set Working Directory}
setwd("/Users/cn/rcode/P3/")
```

```{r Load_the_Parameters}
# Set the date you would like to review
# and all other days will be removed.
report_date <- as.Date(params$dayOfAnalysis)

# Setup variables for the limits for the charts used for
# defining the horizontal bars
voltageHigh = params$voltageHigh
voltageLow  = params$voltageLow

# Power Factor High Low limits used for the plots for
# defining the vertical bars
pfHigh = params$pfHigh
pfLow  = params$pfLow

# Upper limit of voltage.  Any voltage above 140v
# will be considerd invalid and removed from the analysis
voltageMaxLimit = params$voltageMaxLimit
voltageLowLimit = params$voltageLowLimit

```

```{r Functions}
# Trim leading and trailing whitespaces
trim <- function (x) gsub("^\\s+|\\s+$", "", x)

```

```{r Load_the_Data - Raw Dataset Descriptions}
# Load the Data
# read date is in local time
powerfactor_raw <- read.csv("data/backup/pf.csv")
voltage_raw <- read.csv("data/voltage_anon.csv")
substations <- read.csv("substations_list_anon.csv")

# Descriptions of the datasets - 'raw' mean uncleaned and not validated
# voltage_raw  - 15 min Intervals
#   readdate: string in yyyy-mm-dd hh:mm:ss format
#   Voltage : integer
#   substationName: string of full substation name
#   meter   : integer

# powerfactor_raw - Hourly Intervals
#   ReadValue: num  
#   station  : 4 Charcter Desc of substation in this system (SCADA System)
#   name     : Unit of Measure 
#       mega watt 
#       mvars delievered
#       mvars received
#   ReadDate : string in yyyy-mm-dd hh:mm:ss format

# substations, lookup table to map station to substation name
#   station  : 4 Charcter Desc of substation in this system (SCADA System)
#   substationName: string of full substation name
#   prettyName: string of full substation name
```

```{r Remove Duplicates from voltage_raw and powerfactor_raw}
voltage_clean_all <- voltage_raw  %>% 
  group_by(readdate,Voltage, substationName, meter) %>% 
  filter(row_number() == 1)

powerfactor_clean_all <- powerfactor_raw  %>% 
  group_by(ReadValue, station, name, ReadDate) %>% 
  filter(row_number() == 1)
```

```{r Date Formating so we can partition out one day}
# POWER FACTOR
# covert the chr date to POSIX
powerfactor_clean_all$dtReadDate <- parse_date_time(powerfactor_clean_all$ReadDate, 
                                            orders="ymd hms")
# Roll back 15min to accomodate interval ending and hours
powerfactor_clean_all$dtReadDate  <- powerfactor_clean_all$dtReadDate  - minutes(10)

# Add day and hour, so we can group by day or hour
powerfactor_clean_all$dtReadDay  <- floor_date(powerfactor_clean_all$dtReadDate,"day")
powerfactor_clean_all$h <- hour(powerfactor_clean_all$dtReadDate)

# VOLTAGE
# covert the chr date to POSIX
voltage_clean_all$dtReadDate <- parse_date_time(voltage_clean_all$readdate, 
                                            orders="ymd hms")
# Roll back 15min to accomodate interval ending and hours
voltage_clean_all$dtReadDate  <- voltage_clean_all$dtReadDate  - minutes(15)

# Add day and hour, so we can group by day or hour
voltage_clean_all$dtReadDay  <- floor_date(voltage_clean_all$dtReadDate,"day")
voltage_clean_all$h <- hour(voltage_clean_all$dtReadDate)
```

```{r powerfactor_clean_all - Add the full substation name to the dataframe}
# Associate the short substation name in the power factor data with the
# longer substation name in the voltage data.
powerfactor_clean_all <- left_join(powerfactor_clean_all, substations, 
                               by= c("station"))
```

```{r Tidy the data set instructions - Tidy Dataset Descriptions}
# Done - Your data set should include: at least 1,000 observations 
#   (both have more than 1000)
# Done - contain at least one categorical variable (you may create one)
#   Lagging & Leading  - Created
# Done - contain at least 8 different variables
# Voltage Data
#   ReadDate
#   dtReadDate (Shifted back 15min for cleaning)
#   dtReadDay 
#   h (hour)
#   hm (hr:min)
#   Substation
#   Meter
#   Voltage (ReadValue)
#   VoltHalf (View in terms of 120v instead of 240v)
#   voltage.bucket
#     (  0, 114],  <= 114
#     (114, 126]   > 114 and <= 126
#     (126, 140]   > 126 and <= 140
#
# Power Factor Data
#   substation name
#   ReadDate
#   dtReadDate (Shifted back 10min for cleaning)
#   dtReadDay
#   h (hour)
#   mw (mega-watts)
#   mvar.delivered
#   mvar.received
#   mvars (mvar.delivered - mvar.received)
#   mwsquared
#   mvarsquared
#   desc (Lagging or Leading)
#   hm (hour:min)
#   pfChart Shifted Value for Leading using formula listed below
#
# Be in a tidy format1 (you may need to clean and reshape the data)
# Column headers are values, not variable names.
# Multiple variables are stored in one column.
# Multiple types of observational units are stored in the same table.
# A single observational unit is stored in multiple tables

```

```{r Power Factor - Pivot the data}
# Pivot the data for the pf calculation
# 'SCADA.Point'name' contains c(megawatts, megavars delievered, megaVars received)
# and our goal is to add each of these into their own column
powerfactor <- spread(powerfactor_clean_all, name, ReadValue)

# rename the columns
# TODO: If the pivot cols are missing code fails
powerfactor <- dplyr::rename(powerfactor, 
                  mw=PMWD3D, 
                  mvar.delivered=PMQD3D, 
                  mvar.received=PMQD3R)
```

```{r Power Factor Date Formatting}
# Date Formating
# The data is interval ending. This means the dates need to be shifted 15 
# minutes back.  Typically for 15 min intervals this is the bucket:
# (1/1/2016 00:15, 00:30, 00:45, 1:00) for hour one.
# (1/1/2016 23:15, 23:30, 23:45, 1/2/2016 00:00) for hour 24.
# Subtracting 15mins from the date will only work for intervals < than one hour
# Solved using lubridate, as.Date and POSXt was a pain.
powerfactor$dtReadDate <- parse_date_time(powerfactor$ReadDate, orders="ymd hms")
powerfactor$dtReadDate <- powerfactor$dtReadDate - minutes(10)
powerfactor$dtReadDay  <- floor_date(powerfactor$dtReadDate,"day")
powerfactor$h <- hour(powerfactor$dtReadDate)
# add back the 15min to match the 15min intervals for voltage
powerfactor$hm <- format(as.POSIXct(powerfactor$dtReadDate + minutes(15), 
                           format="%Y-%m-%d %H:%M"),
                           format="%H:%M")

# Convert hour numeric, we will do the same for v_dedupe
# so we can join these columns
powerfactor$h <- as.numeric(powerfactor$h)
```

```{r Power Factor calcualte MVARs}
# Calculate the total MVARs which is the total used 
# (mvar.delivered-mvar.received)
# mvar.received can be .na, when this occurs just use the mvar.delievered
powerfactor$mvar <- ifelse(!is.na(powerfactor$mvar.received), 
                       (powerfactor$mvar.delivered-powerfactor$mvar.received) 
                       ,powerfactor$mvar.delivered)

# Add the leading and lagging field to pf_wide.
# If mvar (+) Lagging, If mvar (-) Leading
powerfactor$desc <- ifelse(powerfactor$mvar < 0, 'Leading' , 'Lagging')

# factor this field
powerfactor$desc <- factor(powerfactor$desc) 
```

```{r Power Factor - Calculate Power Factor}
# Calculate the power factor using formuals below
# Power Factor = kw / (kw^2 + (mvar.delivered-mvar.received)^2)  
# Power Factor = mw / (mw^2 + (mvar)^2)  
powerfactor$pf <- powerfactor$mw / 
    sqrt((powerfactor$mw^2 + powerfactor$mvar^2))

```

```{r Power Factor Right Shift (pf) for lagging and leading}
# Right shift the leading power factors for the plotting
# Chart center is 1, and to show the direction of the 
# phase angle we shift values to the right for the leading power factor.
# To right shift ~ (1-pf) + 1
powerfactor$pfChart <- ifelse( powerfactor$mvar < 0, 
                               (1-powerfactor$pf)+1 ,
                               powerfactor$pf)

```

```{r Power Factor add mw and mvar squared for future calculations}
# Right shift the leading power factors for the plotting
# Chart center is 1, and to show the direction of the 
# phase angle we shift values to the right for the leading power factor.
powerfactor$mwsquared <- powerfactor$mw^2
powerfactor$mvarsquared <- powerfactor$mvar^2

```

```{r Power Factor - Map which hours are Lagging and Leading in the dataframe}
  #hours
  #maybe some other value too... which I can do stats on... 
```

```{r Power Factor Trim Substation Name}
# Trim and Upper the Substation Name to scrub the data, some text may have 
# white space around the characters
powerfactor$substationName <- trim(powerfactor$substationName)
powerfactor$substationName <- toupper(powerfactor$substationName)

```

```{r TIDY - Power Factor chose columns to keep}
# Create new dataframes with just the data we need to review
# tidy up
powerfactor_tidy <- powerfactor[c("substationName",
                              "ReadDate",
                              "dtReadDate",
                              "dtReadDay",
                              "h", 
                              "hm", 
                              "mw",
                              "mvar.delivered",
                              "mvar.received",
                              "mvar", 
                              "mwsquared",
                              "mvarsquared",
                              "pf", 
                              "pfChart", 
                              "desc")]
# sort the data
powerfactor_tidy <- arrange(powerfactor_tidy, substationName, dtReadDate)  

```

```{r powerfactor_tidy add a cut for the pf ranges}
# When bucketing this data, I think all we really care
# about is the actual range without the direction since
# we are interested in finding the most intervals in 
# the narrow range , (.98 to 1 to .98), so we can
# ignore pfChart and the direction
# we want the most of our values in the 0.98, 1 range
powerfactor_tidy$pf.range <-cut(powerfactor_tidy$pf, 
              breaks=c(0,.88,.90,.92,.94,.96,.98,1),
              labels=c("(0,0.88.]",
                       "(0.88,0.90]", 
                       "(0.90,0.92]", 
                       "(0.92,0.94]",
                       "(0.94,0.96]",
                       "(0.96,0.98]",
                       "(0.98,1]"))
              
```

```{r Votlage Date Formatting}
voltage_clean_all$dtReadDate <- parse_date_time(voltage_clean_all$readdate, 
                                            orders="ymd hms")

# Roll back 15min to accomodate interval ending and hours
# The data is interval ending. This means the dates need to be shifted 15 
# minutes back.  Typically for 15 min intervals this is the bucket:
# (1/1/2016 00:15, 00:30, 00:45, 1:00) for hour one.
# (1/1/2016 23:15, 23:30, 23:45, 1/2/2016 00:00) for hour 24.
voltage_clean_all$dtReadDate  <- voltage_clean_all$dtReadDate  - minutes(15)

# Add day and hour, so we can group by day or hour
voltage_clean_all$dtReadDay  <- floor_date(voltage_clean_all$dtReadDate,"day")
voltage_clean_all$h <- hour(voltage_clean_all$dtReadDate)

# Create the hour:min 00:00 
# RFC3339     = "2006-01-02T15:04:05Z07:00"
voltage_clean_all$hm <- format(as.POSIXct(voltage_clean_all$readdate, 
                           format="%Y-%m-%d %H:%M"),
                           format="%H:%M")

# force the value to numeric like we have for powerfactor$h
voltage_clean_all$h <- as.numeric(voltage_clean_all$h)
```

```{r Voltage create new variable from 240v to 120v}
# Divide all the votlages by 2 since we have 240v want to see 120v instead
# requirement: from customer
# Save original value in new column
voltage_clean_all$VoltsHalf <- voltage_clean_all$Voltage / 2

```

```{r Voltage Remove Spikes - values over the voltageMaxLimit}
# Remove bad values over the voltage max limit
voltage_clean_all <- subset(voltage_clean_all, 
                        voltage_clean_all$VoltsHalf < voltageMaxLimit)
# Remove all values under the voltageLowLimit
voltage_clean_all <- subset(voltage_clean_all, 
                        voltage_clean_all$VoltsHalf > voltageLowLimit)
```

```{r Voltage Trim Substation Name}
# Trim and Upper the Substation Name for Consistency
voltage_clean_all$substationName <- trim(voltage_clean_all$substationName)
voltage_clean_all$substationName <- toupper(voltage_clean_all$substationName)

```

```{r Voltage Create Buckets}
# Create the buckets for the voltage
# (  0, 114],  <= 114
# (114, 126]   > 114 and <= 126
# (126, 140]   > 126 and <= 140
voltage_clean_all$voltage.bucket <- cut(voltage_clean_all$VoltsHalf,
                                    breaks = c(0,voltageLow,120,voltageHigh,140))#,
                                    #labels=c('Low','Med', 'Best','High'))
```

```{r TIDY - Voltage dataframes - chose columns to keep}
# Create new dataframes with just the data we need to review
# tidy up
voltage_tidy <- voltage_clean_all[c("substationName",
                          "meter",
                          "readdate", 
                          "dtReadDate",
                          "dtReadDay",
                          "h", 
                          "hm", 
                          "Voltage",
                          "VoltsHalf",
                          "voltage.bucket")]

# meter is a factor
voltage_tidy$meter <- factor(voltage_tidy$meter)

# sort the data
voltage_tidy <- arrange(voltage_tidy, substationName, meter, dtReadDate)  
```










```{r DELETE ALL DAYS EXCEPT ONE DAY, include=FALSE}
# THIS CAN BE DONE LAST ... OR IN THE INDIVIDUAL PLOTS
# drop all dates but one day, uses the d(date) parameter
powerfactor_raw_oneday <-  subset(powerfactor_tidy, 
                                  as.Date(powerfactor_tidy$dtReadDay) 
                                  == as.Date(report_date))

# drop/remove all dates but one day, using the report_data variable
# for this first analaysis, uses the d(date) parameter
voltage_raw_oneday <-  subset(voltage_tidy, as.Date(voltage_tidy$dtReadDay) 
                       == as.Date(report_date))
```







```{r Summarise voltage_tidy to the hour - summary_voltage}
# Voltage has 96 intervals per day and power factor 24 intervals per day
# Rollup the Voltage to the hour by day

# Power Factor could be lagging or leading per hour and this is the lowest
# time interval we can support.  To merge this data with the voltage
# data I will have to first summarise the voltage data to the hour
# adding in the meter counts
summary_voltage <- voltage_tidy %>%
  group_by(substationName, dtReadDay, h)  %>%
  summarise( v.min = min(VoltsHalf),
             v.mean = mean(VoltsHalf),
             v.median = median(as.numeric(VoltsHalf)),
             v.std.dev = sd(VoltsHalf),
             v.varience = var(VoltsHalf),
             v.max = max(VoltsHalf), 
             v.sem = sd(VoltsHalf)/sqrt(length(VoltsHalf)),
             v.meters = length(unique(meter)),
             v.intervals = n())  %>% 
  arrange(substationName, dtReadDay, h)
```

```{r Create summary_total from Merging powerfactor_tidy and summary_voltage to the hour}
# summary_total contains data aggregated to the hour
# summary voltage data is aggreagted from 15min to the hour
# leaving off the meters
summary_total <- left_join(summary_voltage, powerfactor_tidy, 
                   by= c("substationName", "dtReadDay", "h"))


summary_total$isLag <- ifelse(summary_total$desc == "Lagging",1,0)
summary_total$isLead <- ifelse(summary_total$desc == "Leading",1,0)
```

```{r Summarise summary_total to the day by desc - summary_total_day}
# Aggregate the data to the day for each substation
# Add count of the number of hours leading and lagging
# Examining the lagging and leading Power Factor to the day is 
# 'good enougth' for this analysis since we are losing fidelity.
# and power factor is typical measure as close to real time as 
# possible.
# We can determine if the day had any leading or lagging values
# when the pfc.values do not match the pf.values, this means
# at least one hour for the day was Leading or Lagging... 

summary_total_day <- summary_total %>%
  group_by(substationName, dtReadDay)  %>%
  summarise( v.min = min(v.min),
             v.mean = mean(v.mean),
             v.median = median(v.median),
             v.max = max(v.max), 
             v.intervals   = sum(v.intervals),
             v.meters = max(v.meters),
             pf.min = min(pf),            
             pf.mean = mean(pf),
             pf.median = median(pf),
             pf.max = max(pf), 
             mw = sum(mw),
             mvar = sum(mvar),
             pfc.min = min(pfChart),            
             pfc.mean = mean(pfChart),
             pfc.median = median(pfChart),
             pfc.max = max(pfChart),
             pf.lag.total = sum(isLag),
             pf.lead.total = sum(isLead)
             
             ) %>% 
  arrange(substationName, dtReadDay)

```

```{r summary_total_day - Add Lagging and Leading Description }
# Based on Total mvar for the day, mark record as lagging or leading
# leading mvar < 0
# lagging mvar > 0, this means power is rolling back to the grid
summary_total_day$desc <- ifelse(summary_total_day$mvar < 0, 
                                 'Leading', 
                                 'Lagging')
```

```{r summary_total_day - Add ranges }
# Calculate the ranges for voltage and power factor
# the idea is to see if the voltage range is lower
# and narrower as well when the power factor has a
# narrower range and maintains a close approximation
# to one when either lagging or leading
# TODO I think this will not work, and what I need
# is to cut with a range of pfChart values or pf values
# and indicator or lagging and leading
summary_total_day$v.range <- summary_total_day$v.max - summary_total_day$v.min
summary_total_day$pf.range <- summary_total_day$pfc.max - summary_total_day$pf.min

```

```{r Power Factor Leading and Lagging Calculation}
# Leading and Lagging Analysis
# Divide the lagging and leading dataframes into individual 
# data frames to help the user review the data in tables
lagging <- subset(summary_total_day, summary_total_day$mvar >  0)
leading <- subset(summary_total_day, summary_total_day$mvar <  0)

# Lagging is to the left
lagging_outliers <- subset(lagging, 
                   lagging$pfc.median < pfLow |
                   lagging$v.median  < voltageLow  |
                   lagging$v.median  > voltageHigh )

# Leading is to the right for the chart
leading_outliers <- subset(leading, 
                    leading$pfc.median < pfLow |
                    leading$pfc.median > (1-pfLow)+1 |
                    leading$v.median < voltageLow |
                    leading$v.median > voltageHigh) 
```

# Univariate Plots Section

## Voltage

Range of days: `r range(voltage_tidy$dtReadDay)`

Summary of Voltage: `r summary(voltage_tidy$VoltsHalf)`

```{r Univariate_Plots Histogram Voltage}
summary(voltage_tidy$VoltsHalf)

ggplot(data = voltage_tidy,aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Histogram of Voltage (bin=0.5)") +
  xlab("voltage")

ggplot(data = voltage_tidy,aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  scale_x_continuous(limits = c(voltageLowLimit, voltageMaxLimit)) +
  theme_bw(base_family= 'Helvetica') +
  ggtitle(paste("Histogram of Voltage (bin=0.5)\nLow:",
                as.character(voltageLowLimit),
                "v High:",
                as.character(voltageMaxLimit),
                " v")) +
  xlab("voltage")

# Facet By voltage.bucket
ggplot(data = voltage_tidy,aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  facet_wrap(~voltage.bucket, ncol = 2, scales = "free") +
  ggtitle("Histogram of Voltage by Bucket (bin=0.5)") +
  xlab("voltage")

```

```{r Univariate_Plots Voltage Facet Day and Substation, fig.width=12, fig.height=12}
ggplot(data = voltage_tidy,aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  facet_wrap(~dtReadDay, ncol = 4) +
  ggtitle("Histogram of Voltage By Day (bin=0.5)") +
  xlab("voltage")

# ggplot(data = voltage_tidy,aes(x = VoltsHalf)) + 
#   geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
#   theme_bw(base_family= 'Helvetica') +
#   facet_wrap(~substationName, ncol = 4) +
#   ggtitle("Histogram of Voltage By Substation (bin=0.5)") +
#   xlab("voltage")

```

```{r Univariate_Plots Voltage Facet hour and substation lenght calc}
height = length(unique(voltage_tidy$substationName))
height = height / 4
```

```{r Univariate_Plots Voltage Hist Facet Substation free scales, fig.width=12, fig.height=len}
ggplot(data = voltage_tidy,aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  facet_wrap(~substationName, ncol = 4, scales = 'free') +
  ggtitle(paste("Histogram of Voltage By Substation (bin=0.5)\n",
                as.character(length(unique(voltage_tidy$dtReadDay))),
                "days combined")) +
  xlab("voltage")
```

## Power Factor
```{r Univariate_Plots Power Factor Histogram}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ]
            ,aes(x = pfChart, fill=desc)) + 
  geom_histogram(binwidth=0.002, color = 'black') + 
  theme_bw(base_family= 'Helvetica') + 
  theme(legend.position="bottom") + 
  theme(legend.title=element_blank())+
  scale_x_continuous(limits = c(0.92, 1.08),
                       breaks=c(.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c("0.92",".94", ".96", ".98","1", ".98", ".96",".94","0.92")) +
  xlab("power factor") +
  ggtitle(paste("Power Factor Histogram (bin=0.002)\n",
                as.character(length(unique(voltage_tidy$dtReadDay))),
                "days combined")) 
```

```{r Univariate_Plots Power Factor Facet Hour and Substation, fig.width=12, fig.height=10}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ]
            ,aes(x = pfChart, fill=desc)) + 
  geom_histogram(binwidth=0.002, color = 'black') +
  facet_wrap(~dtReadDay, ncol=4) +
  theme_bw(base_family= 'Helvetica') + 
  theme(legend.position="bottom") + 
  theme(legend.title=element_blank())+
  scale_x_continuous(limits = c(0.92, 1.08),
                       breaks=c(.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c("0.92",".94", ".96", ".98","1", ".98", ".96",".94","0.92")) +
  xlab("power factor") +
  ggtitle("Power Factor Histogram By Hour (bin=0.002)")
  

ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ]
            ,aes(x = pfChart, fill=desc)) + 
  geom_histogram(binwidth=0.002, color = 'black') +
  facet_wrap(~substationName, ncol=3, scales = "free_y") +
  theme_bw(base_family= 'Helvetica') + 
  theme(legend.position="bottom") + 
  theme(legend.title=element_blank())+
  scale_x_continuous(limits = c(0.92, 1.08),
                       breaks=c(.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c("0.92",".94", ".96", ".98","1", ".98", ".96",".94","0.92")) +
  xlab("power factor") +
  ggtitle(paste("Power Factor Histogram By Substation (bin=0.002)\n",
                as.character(length(unique(voltage_tidy$dtReadDay))),
                "days combined")) 
```

### MegaWatts, MVARs, and MVARS Squared

```{r Univariate_Plots Mvar mvarsquared mw  }
ggplot(data = powerfactor_tidy,aes(x = mw)) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Histogram of MW (bin=0.25)") +
  xlab("MW")

summary(powerfactor_tidy$mw)

ggplot(data = powerfactor_tidy,aes(x = mvar)) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Histogram of MVAR (bin=0.25)") +
  xlab("MVAR")

summary(powerfactor_tidy$mvar)

ggplot(data = powerfactor_tidy,aes(x = mvarsquared)) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Histogram of MVAR^2 (bin=0.25)") +
  xlab("mvarsquared")

summary(powerfactor_tidy$mvarsquared)

```

```{r Univariate_Plots MW Facet Hour and Substation, fig.width=12, fig.height=12}
ggplot(data = powerfactor_tidy,aes(x = mw)) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  facet_wrap(~substationName, ncol=3) +
  ggtitle("Histogram of MW by Substation (bin=0.25)") +
  xlab("MW")

ggplot(data = powerfactor_tidy,aes(x = mw)) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  facet_wrap(~dtReadDay, ncol=4) +
  ggtitle("Histogram of MW By Day (bin=0.25)") +
  xlab("MW")

```

# Univariate Analysis
### What is the structure of your dataset?

### Voltage Dataset
The voltage dataset contains endpoints collecting readings
every 15 minutes.

```{r Structure of voltage_tidy}
str(voltage_tidy,give.attr = FALSE)
```

### Power Factor Dataset
The power factor dataset contains endpoints collecting readings
every 60 minutes.

```{r Structure of powerfactor_tidy}
str(powerfactor_tidy)
```

### Substation Dataset
Used for mapping the Voltage and Power Factor datasets together.

```{r Structure of substations}
str(substations)
```


### Structure of Voltage Data - uncleaned
```{r Structure - Voltage Original}
str(voltage_raw)
```

### Structure of Power Factor - uncleaned
```{r Structure - Power Factor Original}
str(powerfactor_raw)
```

```{r Remove Voltage Staging Tables}
rm(voltage_raw, voltage_clean_all)
```

```{r Remove Power Factor Staging Tables}
rm(powerfactor_raw, powerfactor_clean_all, powerfactor)
```

### What is/are the main feature(s) of interest in your dataset?
Voltage and Power Factor by time interval for each deliver point (substation)
are the main features. One goal is to determine if we can predict power factor.

### What other features in the dataset do you think will help support your
investigation into your feature(s) of interest?
* mega watt  
* mega var delivered  
* mega var received  

### Did you create any new variables from existing variables in the dataset?
I created the power factor, mvars and the direction of the power factor
(lagging and leading).

New variables where created for these datasets:

#### powerfactor_tidy and voltage_tidy
* dtReadDate
* dtReadDay
* h
* hm

#### voltage_tidy
* voltage.bucket
* VoltsHalf

#### powerfactor_tidy
* mvars
* mvarsquared
* Power Factor
* Power Factor Chart (Shifted Value)
* desc - the direction of the power factor (Lagging vs Leading)


### Of the features you investigated, were there any unusual distributions? 

Austin and Dallas have bimodel distibutions of voltage.

```{r willamsport and mtpleasant hist , fig.width=12, fig.height=6}
austin <-
  ggplot(data = subset(voltage_tidy, substationName=="AUSTIN"),aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("AUSTIN (bin=0.5)") +
  xlab("voltage")

dallas <-ggplot(data = subset(voltage_tidy, substationName=="DALLAS"),aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("DALLAS (bin=0.5)") +
  xlab("voltage")

grid.arrange(austin, dallas, ncol=2)

```

The long leading tail on the voltage histogram, has a larger range in the data in the lower range than in the upper range.

```{r lower and upper range of voltage hist, fig.width=12, fig.height=6}
lower <-
  ggplot(data = subset(voltage_tidy, VoltsHalf < 117),aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage < 117v (bin=0.5)") +
  xlab("voltage")

upper <-
  ggplot(data = subset(voltage_tidy, VoltsHalf > 126),aes(x = VoltsHalf)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage > 126v (bin=0.5)") +
  xlab("voltage")

grid.arrange(lower, upper, ncol=2)

lowcount = count(subset(voltage_tidy, VoltsHalf < 117,select = c(VoltsHalf)))
lowrange = range(subset(voltage_tidy, VoltsHalf < 117,select = c(VoltsHalf)))

highcount = count(subset(voltage_tidy, VoltsHalf > 126,select = c(VoltsHalf)))
highrange = range(subset(voltage_tidy, VoltsHalf > 126,select = c(VoltsHalf)))

```

Number of voltage intervals < 117: `r count(subset(voltage_tidy, VoltsHalf < 117,select = c(VoltsHalf)))`  
The range is : `r range(subset(voltage_tidy, VoltsHalf < 117,select = c(VoltsHalf)))`

Number of voltage intervals > 126: `r count(subset(voltage_tidy, VoltsHalf > 126,select = c(VoltsHalf)))`  
The range is : `r range(subset(voltage_tidy, VoltsHalf > 126,select = c(VoltsHalf)))`

### Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Various methods were used to clean the data.  For instance the ReadDates for
the voltage intervals are in ending interval.  The interval starts at 3/2/2016
00:15min and ends on 3/3/2016 00:00.  To associate the 15 minute intervals 
with the correct hour and day, we had to roll back each 15 minute interval 
by 15 minutes.

The power factor data needed to be pivoted to get the data into a tidy format 
as well. The orignal data has the MegaWatt, MegaVars Delivered and Received 
in the same column, these values were split out into their own columns.

The substation names can have leading and trailing spaces so this data needed 
to be trimed.

# Bivariate Plots Section
### Voltage Plots
#### Box Plots
```{r Bivariate_Plots Boxplot Voltage, fig.width=12, fig.height=10}
ggplot(data=voltage_tidy[!is.na(voltage_tidy$VoltsHalf), ], 
       aes(factor(substationName), VoltsHalf)) +
    geom_boxplot(outlier.colour = "red") + 
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    geom_hline(yintercept = voltageHigh, color='blue') +
    geom_hline(yintercept = voltageLow, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Voltage by Substation" ) 

ggplot(data=voltage_tidy[!is.na(voltage_tidy$VoltsHalf), ], 
       aes(factor(substationName), VoltsHalf)) +
    geom_boxplot(outlier.colour = "red") + 
    facet_wrap(~ dtReadDay) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    geom_hline(yintercept = voltageHigh, color='blue') +
    geom_hline(yintercept = voltageLow, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Substation Voltage by Day" ) 

ggplot(data=voltage_tidy[!is.na(voltage_tidy$VoltsHalf), ], 
       aes(x=factor(h), y=VoltsHalf)) +
    geom_boxplot(outlier.colour = "red") + 
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    geom_hline(yintercept = voltageHigh, color='blue') +
    geom_hline(yintercept = voltageLow, color='blue') +
    ggtitle("Boxplot of Voltage by hour" ) 
```

```{r Bivariate_Plots Boxplot Voltage by Hour Facet Substations, fig.width=12, fig.height=16}
ggplot(data=voltage_tidy[!is.na(voltage_tidy$VoltsHalf), ], 
       aes(x=factor(h), y=VoltsHalf)) +
    geom_boxplot(outlier.colour = "red") + 
    facet_wrap(~substationName, ncol = 3) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    geom_hline(yintercept = voltageHigh, color='blue') +
    geom_hline(yintercept = voltageLow, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Substation Voltage by hour " ) 

```

#### Scatter Plots
```{r Voltage 15 Minute Intervals By Hour, fig.width=12, fig.height=6}
# TODO Add Low and High Value variables on the page or just make y axis 
# show the points for the hline
# TO MUCH FOR R
ggplot(data=voltage_tidy, aes(x=hm, y=VoltsHalf)) +
  geom_point(aes(color=substationName),
             position=position_jitter(h=1/4),
             alpha=1/2, 
             size=1) + 
  scale_y_continuous(limits = c(100, 130)) +
  scale_x_discrete(breaks=c("00:00","06:00","12:00","18:00","23:00")) +
  scale_fill_brewer(palette="Set2") +
  geom_hline(yintercept=voltageLow) +
  geom_hline(yintercept=voltageHigh) +
  xlab("hr:min") +
  ylab("voltage") +
  #ggtitle(paste("Voltage 15min Intervals", as.character(voltageLow)," ")) +
  theme_bw(base_family= 'Helvetica') +
  theme(axis.text.x = element_text(angle = -90, hjust = 0))  +
  theme(legend.position="bottom")

```

##### Dallas Substation Daily Voltage By Hour
```{r Voltage Scatter Plot by Substation Facet with alpha, fig.width=10, fig.height=14}

ggplot(data=subset(voltage_tidy,substationName == "DALLAS"), aes(x=hm, y=VoltsHalf)) +
  geom_point(alpha=1/5) + 
  facet_wrap(~ dtReadDay, ncol = 3) +
  scale_x_discrete(breaks=c("00:00","06:00","12:00","18:00","23:00")) +
  xlab("hr:min") +
  ylab("voltage") +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  ggtitle("DALLAS Voltage by Time (hour:min)" )

```

The scatter plot shows the data along the time axis for the intervals for the 
day. The interesting point in this chart, which is similar to the histogram is 
how the shading changes from dark to grey, which is the points stacking on top 
of each other.  

It takes 5 points on top of each other to make a solid point on this chart.  
This demonstrates how the data is spread out over the ranges through the day 
by hour.

We can see there seems to be one metering point with high voltage which is
consistently above the other voltage readings.

```{r Summary per meter}
meter_summary_volts <- subset(voltage_tidy,substationName == "DALLAS") %>%
  group_by(meter, dtReadDay)  %>%
  summarise( v.min = min(VoltsHalf),
             v.mean = mean(VoltsHalf),
             v.median = median(VoltsHalf),
             v.max = max(VoltsHalf), 
             v.intervals   = n()
             ) %>% 
  arrange(meter, dtReadDay)
```

##### Dallas Substation Meters
If we search all the meters on the Dallas substation, and sort descending
by voltage we can obtain a list of meters with the highest voltage.
```{r}
kable(head(meter_summary_volts[order(desc(meter_summary_volts$v.max)),],
           3,digits=2))
```

```{r Find the meter with the issue}
ggplot(data=subset(voltage_tidy,meter == 300686)
       , aes(x=hm, y=VoltsHalf)) +
  geom_point(alpha=1/5) + 
  facet_wrap(~ dtReadDay, ncol = 3) +
  scale_x_discrete(breaks=c("00:00","06:00","12:00","18:00","23:00")) +
  xlab("hr:min") +
  ylab("voltage") +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  ggtitle("Meter 300686 Voltage by Time" )
```

##### Voltage Descriptive Statistics
```{r Voltage Summary}
summary(voltage_tidy$VoltsHalf)
```

###### Standard Deviation
```{r Voltage Standard Deviation}
sd(voltage_tidy$VoltsHalf)
```

###### Range 
```{r Voltage Range}
range(voltage_tidy$VoltsHalf)
```

###### Quantiles
```{r Voltage Quantiles}
quantile(voltage_tidy$VoltsHalf)
```

### Mega Watts
```{r Boxplot Mega Watts by Sub, fig.width=10, fig.height=6}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$mw), ], 
       aes(factor(substationName), mw)) +
    geom_boxplot(outlier.colour = "red") + 
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("MW") + 
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of MW by Substation" )
```

##### Mega Watts Descriptive Statistics
```{r MW Summary}
summary(powerfactor_tidy$mw)
```

### MegaVAR
```{r BoxPlot MegaVar by Sub and Day, fig.width=10, fig.height=6}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$mvar), ], 
       aes(factor(substationName), mvar)) +
    geom_boxplot(outlier.colour = "red") + 
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    geom_hline(yintercept = 0, color='red') +
    xlab("") +
    ylab("MVAR") + 
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of MVAR by Substation" )

ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$mvar), ], 
       aes(factor(substationName), mvar)) +
    geom_boxplot(outlier.colour = "red") + 
    facet_wrap(~dtReadDay) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    geom_hline(yintercept = 0, color='red') +
    xlab("") +
    ylab("MVAR") + 
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of MVAR Substation by Day" )

```

##### MegaVAR Descriptive Statistics
```{r MegaVAR Summary}
summary(powerfactor_tidy$mvar)
```

### System Total MegaWatts, Mega VARs, MegaVars Squared   
```{r Mega Watts By Hour Total for system}
load_hour_totals <- powerfactor_tidy %>%
  group_by(h)  %>%
  summarise(mw.total=sum(mw, na.rm=TRUE),
            mw.mean = mean(mw, na.rm=TRUE),
            mw.median = median(mw, na.rm=TRUE),            
            mw.min = min(mw, na.rm=TRUE),
            mw.max = max(mw, na.rm=TRUE),
            mvar.total=sum(mvar, na.rm=TRUE),
            mvar.mean = mean(mvar, na.rm=TRUE),
            mvar.median = median(mvar, na.rm=TRUE),            
            mvar.min = min(mvar, na.rm=TRUE),
            mvar.max = max(mvar, na.rm=TRUE),  
            mvarsquared.total=sum(mvarsquared, na.rm=TRUE),
            pf.mean = mean(mvar, na.rm=TRUE),
            pf.median = median(mvar, na.rm=TRUE),
            pf.min = min(mvar, na.rm=TRUE),
            pf.max = max(mvar, na.rm=TRUE), 
            n = n())  %>% 
  arrange(h)

ggplot(data=load_hour_totals, aes(x=h, y=mw.total)) +
  geom_line() +
  ggtitle("Mega Watt by Hour - Total System" ) +
  ylab("MW") +
  theme_bw(base_family= 'Helvetica')

ggplot(data=load_hour_totals, aes(x=h, y=mvar.total)) +
  geom_line() +
  ggtitle("Mega VAR by Hour - Total System" ) +
  geom_hline(yintercept = 0, color='red') +
  ylab("MVARs") +
  theme_bw(base_family= 'Helvetica')

ggplot(data=load_hour_totals, aes(x=h, y=mvarsquared.total)) +
  geom_line() +
  ggtitle("(Mega VAR)^2 by Hour - Total System" ) +
  ylab("(Mega VAR)^2") +
  theme_bw(base_family= 'Helvetica')

# ggplot(data=load_hour_totals, aes(x=h, y=pf.min)) +
#   geom_line() +
#   geom_line(data=load_hour_totals, aes(x=h, y=pf.mean)) +
#   geom_line(data=load_hour_totals, aes(x=h, y=pf.median)) +
#   geom_line(data=load_hour_totals, aes(x=h, y=pf.max)) +
#   ggtitle("pf.median" ) +
#   ylab("pf.median") +
#   theme_bw(base_family= 'Helvetica')

```








```{r Mega Watts By Substation By Hour, fig.width=12, fig.height=14}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$mw), ]
       , aes(x=h, y=mw, colour = factor(dtReadDay))) +
  geom_point() + 
  facet_wrap(~substationName, scales = 'free', ncol=3) +
  ggtitle("Mega Watt Scatter Plot by Substation" ) +
  theme_bw(base_family= 'Helvetica') 

#TODO
# To 10 - Meters for each substation, those using the most power (kWs)
# Meters using the highest percent of the entire load
# Meters using the highest load for each substation... 
```
The multiple points are the 11 days we have in this dataset.

### Power Factor Plots

#### Box Plots

```{r Bivariate_Plots Boxplot Power Factor, fig.width=10, fig.height=5}
# Reshape the y axis to fit the power factor use case / user expectations
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(factor(substationName), pfChart)) +
    geom_boxplot(outlier.colour = "red") + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("power factor") + 
    geom_hline(yintercept = 1, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Power Factor by Substation" )

# Reshape the y axis to fit the power factor use case / user expectations
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(factor(h), pfChart)) +
    geom_boxplot(outlier.colour = "red") + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("power factor") + 
    geom_hline(yintercept = 1, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Power Factor by hour" )


```

```{r Boxplot of Power Factor by hour for Substations,fig.width=10, fig.height=15}
# this might need it's own code block
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(factor(h), pfChart)) +
    geom_boxplot(outlier.colour = "red") + 
    facet_wrap(~substationName, ncol = 3) +
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("power factor") + 
    geom_hline(yintercept = 1, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Power Factor by hour for Substations" )
```


#### Scatter Plots
```{r Power Factor By Hour Scatter Plot,fig.width=12, fig.height=6}
ggplot(data=powerfactor_tidy, aes(x=h, y=pfChart)) +
  geom_point(position=position_jitter(w=0.5, h=0), alpha=1/5) + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  #scale_y_continuous(limits = c(100, 130)) +
  #scale_x_discrete(breaks=c("00:00","06:00","12:00","18:00","23:00")) +
  #scale_fill_brewer(palette="Set2") +
  geom_hline(yintercept=1) +
  xlab("hour") +
  ylab("power factor") +
  ggtitle("Power Factor by Hour Scatter") +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.position="bottom")
```

```{r Power Factor By Hour Scatter Plot Facet Day,fig.width=12, fig.height=6}
ggplot(data=powerfactor_tidy, aes(x=h, y=pfChart)) +
  geom_point(position=position_jitter(w=0.5, h=0), alpha=1/5) + 
  scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  facet_wrap(~dtReadDay, ncol = 3) +
  geom_hline(yintercept=1) +
  xlab("hour") +
  ylab("power factor") +
  ggtitle("Power Factor by Hour Scatter Facet Day") +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.position="bottom")
```

```{r Power Factor By Hour Scatter Plot Facet Day Dallas,fig.width=12, fig.height=6}
ggplot(data=subset(powerfactor_tidy, substationName=="DALLAS"), 
       aes(x=h, y=pfChart)) +
  geom_line() + 
  scale_y_continuous(limits = c(0.88, 1.12),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  facet_wrap(~dtReadDay, ncol = 3) +
  geom_hline(yintercept=1) +
  xlab("hour") +
  ylab("power factor") +
  ggtitle("Dallas Power Factor by Hour Facet Day") +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.position="bottom")
```

### Voltage.Min vs [MW, MVAR]
```{r Voltage vs X, fig.width=12, fig.height=6}
v1 <- 
  ggplot(data=summary_total,aes(x=mw, y=v.min)) + 
  geom_point(alpha=1/2, color = 'blue') +
  theme_bw(base_family= 'Helvetica') 

v2 <- 
  ggplot(data=summary_total,aes(x=mvar, y=v.min)) + 
  geom_point(alpha=1/2, color = 'blue') +
  theme_bw(base_family= 'Helvetica') 

grid.arrange(v1, v2, ncol = 2)

```

### Voltage.Mean vs [MW, MVAR, Power Factor]
```{r Voltage vs MW MVAR PowerFactor, fig.width=12, fig.height=4}

# log2, log2, sqrt, 
vm1 <- 
  ggplot(data=summary_total,aes(x=mw, y=v.mean)) + 
  geom_point(alpha=1/2, color = 'green') +
  
  #scale_x_sqrt() +
  #cale_y_sqrt() +
  theme_bw(base_family= 'Helvetica') 

vm2 <- 
  ggplot(data=summary_total,aes(x=mvar, y=v.mean)) + 
  geom_point(alpha=1/2, color = 'green') +
  theme_bw(base_family= 'Helvetica') 

vm3 <- 
  ggplot(data=summary_total,aes(x=pfChart, y=v.mean)) + 
  geom_point(alpha=1/2, color = 'green')  +
  theme_bw(base_family= 'Helvetica') 

grid.arrange(vm1, vm2, vm3, ncol = 3)

```

### Voltage [Min, Mean, Median, Max] vs MW

```{r Voltage Min Mean Median Max vs MW, fig.width=10, fig.height=10}
vmw1 <- 
  ggplot(data=summary_total,aes(x=v.min, y=mw)) + 
  geom_point(alpha=1/2, color = 'blue')  +
  theme_bw(base_family= 'Helvetica') 

vmw2 <- 
  ggplot(data=summary_total,aes(x=v.mean, y=mw)) + 
  geom_point(alpha=1/2, color = 'blue') +
  theme_bw(base_family= 'Helvetica') 

vmw3 <- 
  ggplot(data=summary_total,aes(x=v.median, y=mw)) + 
  geom_point(alpha=1/2, color = 'blue') +
  theme_bw(base_family= 'Helvetica') 

vmw4 <- 
  ggplot(data=summary_total,aes(x=v.max, y=mw)) + 
  geom_point(alpha=1/2, color = 'blue') +
  theme_bw(base_family= 'Helvetica') 

grid.arrange(vmw1, vmw2, vmw3, vmw4, ncol = 2)

```

```{r Mega Watt vs Power Factor (Shifted Value), fig.width=12, fig.height=10}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pf), ], aes(x=mw, y=pfChart)) +
  geom_point() + 
  stat_smooth(method = "lm") +
  facet_wrap(~substationName, ncol=3, scales = "free_x") +
  scale_y_continuous(limits = c(min(powerfactor_tidy$pfChart), max(powerfactor_tidy$pfChart)),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
  geom_hline(yintercept = 1) +
  ylab("power factor") +
  ggtitle("Mega Watt vs Power Factor" ) +
  theme_bw(base_family= 'Helvetica') 
```


## Additional Substation Analysis

### Substations meters where the voltage is less than nominal
```{r Substations Meters volts less than or equal to 114 volts}

# It would be nice to color the rows for those which need
# to be pointed out and this would require html table
substations_below_114 <- subset(voltage_tidy,VoltsHalf<=114)
substations_below_114 <- select(substations_below_114,
                                substationName,
                                meter,
                                VoltsHalf,
                                readdate,
                                dtReadDay)

# it would be usefull here to have total count of intervals
substation_meter_review_114 <-
  substations_below_114 %>% 
    group_by(substationName, meter, dtReadDay) %>% 
    summarise(count = n())  %>% 
        arrange(count,substationName, dtReadDay)
```

### Voltage Low Limit 
Meters where more than 4 intervals below the `r voltageLow` volts threshold.

#### Sample of Meters 
```{r Voltage by Hour for By Meter less than 114 having, fig.width=12, fig.height=10}
# filter the list for meters where count > 4
# would be helpful to know the subtation
# this needs to be reworked because we have multiple days
# need to sort by the lowest offenders and then pick
# 9 meters to show...

meters114 <- subset(substation_meter_review_114, 
                    count>4)
meters <- voltage_tidy[(voltage_tidy$meter %in% meters114$meter),]



ggplot(data=subset(meters, substationName=='HIGHLAND PARK' & meter == 301146), 
       aes(x=hm, y=VoltsHalf, color=factor(dtReadDay))) +
  geom_point() + 
  facet_wrap(~ meter, ncol = 3) +
  scale_x_discrete(breaks=c("00:00","06:00","12:00","18:00","23:00")) +
  xlab("hr:min") +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  ggtitle("Austin Voltage by hour, Less than nominal meters" ) +
  geom_hline(yintercept = 114, color='red') +
  theme(legend.position="bottom") +
  theme(legend.title=element_blank()) +
  geom_text(data = meters, aes(15, 100, label = substationName), size=3)

```

#### Top Top 10 Substations Meter Counts for Low Voltage Meters
```{r}
# SORT FOR DISPLAY THIS IS LONGER THAN 85
substation_meter_review_114 <- substation_meter_review_114[order(-substation_meter_review_114$count),]
kable(head(substation_meter_review_114,10), digits=2, title="Meters where voltage < 114")
```


### Voltage High Limit
Meters where more than 4 intervals below the `r voltageHigh` volts threshold.

```{r Voltage by Hour for By Meter greater than 128, fig.width=12, fig.height=24}
# TODO how can I dynamically size the height based on # of meters
substations_above_128 <- subset(voltage_tidy,VoltsHalf>126)
substations_above_128 <- select(substations_above_128,
                                substationName,
                                meter,
                                VoltsHalf,
                                readdate,
                                dtReadDay)

substation_meter_review_128 <-
  substations_above_128 %>% 
    group_by(substationName, meter, dtReadDay) %>% 
    summarise(count = n())  %>% 
        arrange(count,substationName, meter, dtReadDay)

meters_128 <- subset(substation_meter_review_128, count>20)
meters_intervals_128 <- voltage_tidy[(voltage_tidy$meter %in% meters_128$meter),]

ggplot(data=meters_intervals_128, aes(x=hm, y=VoltsHalf, color=substationName)) +
  geom_point(alpha=1/1) + 
  facet_wrap(~ meter, ncol = 3) +
  scale_x_discrete(breaks=c("00:00","06:00","12:00","18:00","23:00")) +
  xlab("hr:min") +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  ggtitle("Voltage by hour (hour:min), Less than nominal meters" ) +
  geom_hline(yintercept = 126, color='red') +
  theme(legend.position="bottom") +
  theme(legend.title=element_blank()) +
  geom_text(data = meters, aes(15, 120, label = substationName), size=3)


```

#### Top Top 10 Substations Meter Counts for High Voltage Meters

```{r substation_meter_review_128 table data}

kable(head(substation_meter_review_128,10), digits=2)
#TODO How do we determine if the one interval or group of them are significant
# so if I have one value that is below 114 or above 126, should I pay attention
# to this value or just ignore it...
```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

#### Features of Interest
* Voltage
  * By Substation
  * By Hour

* Power Factor
  * By Substation
  * By Hour

#### Other Features
* MW
* MVAR

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

This shows an interesting trend starting at 11AM (11:00 hours) until 
10PM (20:00 hrs). The power factor spreads over a wider range.  This is
interesting on a system wide review, however we are more concerned with
the power factor for each delivery point. 

```{r Power Factor vs Hour}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(factor(h), pfChart)) +
    geom_boxplot(outlier.colour = "red") + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("power factor") + 
    geom_hline(yintercept = 1, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Power Factor by hour" )

```

### What was the strongest relationship you found?
The strongest relationship I found is between megawatts and power factor.  
As the megawatts increases the power factor approaches the 1, for each
of the Substations for this single day investigation.

```{r Mega Watt vs Power Factor, fig.width=12, fig.height=10}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pf), ], aes(x=mw, y=pfChart)) +
  geom_point() + 
  stat_smooth(method = "lm") +
  facet_wrap(~substationName, ncol=3, scales = "free_x") +
  scale_y_continuous(limits = c(min(powerfactor_tidy$pfChart), max(powerfactor_tidy$pfChart)),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
  geom_hline(yintercept = 1) +
  ggtitle("Mega Watt vs Power Factor (Shifted Value)" ) +
  ylab("power factor") +
  theme_bw(base_family= 'Helvetica') 
```

# Multivariate Plots Section

```{r Mega Watt vs Power Factor with Linear Model, fig.width=10, fig.height=6}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pf), ], 
       aes(x=mw, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)  +
  ggtitle("Mega Watt vs Power Factor" ) +
  ylab("power factor") +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank())
```

```{r Voltage vs Power Factor with Linear Model, fig.width=10, fig.height=6}
ggplot(data=summary_total[!is.na(summary_total$pfChart), ], 
       aes(x=v.mean, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)   +
  scale_y_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +   
  ggtitle("Voltage vs Power Factor" ) +
  ylab("power factor") +
  xlab("hourly avg. voltage") +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank())

ggplot(data=summary_total[!is.na(summary_total$pfChart), ], 
       aes(x=v.mean, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  facet_wrap(~substationName) +
  geom_smooth(method=lm)   +
  scale_y_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +   
  ggtitle("Voltage vs Power Factor" ) +
  ylab("power factor") +
  xlab("hourly avg. voltage") +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank())
```

### Correlation Analysis
Coefficient, r  

Strength of Association  |  Positive   |  Negative    
-------------------------|-------------|--------------
Small                    | .1 to .3    | -0.1 to -0.3 
Medium                   | .3 to .5    | -0.3 to -0.5  
Large                    | .5 to 1.0   | -0.5 to -1.0 


#### Correlation Matrix
Review the correlation between power factor and voltage using Pearsons.
```{r Correlation Matrix}
  
cor_matrix <- 
  round(cor(powerfactor_tidy[c("pf",
                       "pfChart",
                       "mw", 
                       "mvar",
                       "mwsquared",
                       "mvarsquared")],
      use="complete.obs",
      method="pearson"
      ) ,4)

```

#### Correlation Plots Power Factor
```{r Correlation Plots for Power Factor, fig.width=10, fig.height=6}
# High to Low
# 0.2938489 pf & mvar
# -0.4169005 pf & mvarsquared
cor0 <- 
  ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pf), ], 
       aes(x=mvar, y=pf)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm, formula = y ~ x)  +
  ggtitle("MVAR vs Power Factor" ) +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank()) +
  annotate("text", label = paste("Cor:",as.character(cor_matrix[4,1:1])), 
           x=Inf, y = Inf, vjust=2, hjust=1, 
           size = 3, 
           colour = "blue")

cor1 <- 
  ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pf), ], 
       aes(x=mvarsquared, y=pf)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)  +
  ggtitle("MVAR^2 vs Power Factor" ) +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title=element_blank())  +
  annotate("text", label = paste("Cor:",as.character(cor_matrix[6,1:1])), 
           x=Inf, y = Inf, vjust=2, hjust=1, 
           size = 3, 
           colour = "blue")

grid.arrange(cor0, cor1, ncol=1)

# High to Low
# pfChart & mw
# pfChart & mvar
# pfChart & mwsquared

```

#### Correlation Plots Power Factor Shifted Value

```{r Correlation Plots for Power Factor Shifted, fig.width=10, fig.height=12}
# High to Low
# pfChart & mw
# pfChart & mvar
# pfChart & mwsquared
cor2 <- 
  ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(x=mw, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)  +
  ggtitle("MW vs Power Factor" ) +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank()) +
    annotate("text", label = paste("Cor:",as.character(cor_matrix[3,2])), 
           x=Inf, y = Inf, vjust=2, hjust=1, 
           size = 3, 
           colour = "blue")

cor3 <- 
  ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(x=mvar, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)  +
  ggtitle("MVAR vs Power Factor" ) +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title=element_blank()) +
  annotate("text", label = paste("Cor:",as.character(cor_matrix[4,2])), 
           x=Inf, y = Inf, vjust=2, hjust=1, 
           size = 3, 
           colour = "blue") 

cor4 <- 
  ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], 
       aes(x=mwsquared, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)  +
  ggtitle("MW^2 vs Power Factor" ) +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title=element_blank()) +
  annotate("text", label = paste("Cor:",as.character(cor_matrix[5,2])), 
           x=Inf, y = Inf, vjust=2, hjust=1, 
           size = 3, 
           colour = "blue") 

grid.arrange(cor2, cor3, cor4, ncol=1)

```

#### Correlation of Voltage & Power Factor
Review the correlation between power factor and voltage using Pearsons.
```{r Correlation between Voltage and Power Factor}
cor.test(summary_total$v.mean, summary_total$pf, method="pearson")

```



### Voltage vs PF with Linear Model

```{r Voltage vs PF with Linear Model Facet Substation, fig.width=10, fig.height=6}
ggplot(data=summary_total[!is.na(summary_total$pfChart), ], 
       aes(x=v.mean, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  geom_smooth(method=lm)  +
  facet_wrap(~substationName, ncol = 3) +
  scale_y_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +   
  ggtitle("Voltage vs Power Factor" ) +
  ylab("power factor") +
  xlab("hourly avg. voltage") +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank())
```

### MegaWatt vs PF with Linear Model by Substation

```{r Multivariate_Plots Facet Substation,fig.width=12, fig.height=10}
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$pfChart), ], aes(x=mw, y=pfChart)) +
  geom_point(aes(color=desc)) + 
  stat_smooth(method = "lm") +
  facet_wrap(~substationName, ncol=3, scales = "free") +
  #scale_x_continuous(limits = c(min(powerfactor_tidy$mw), max(powerfactor_tidy$mw))) + 
  scale_y_continuous(limits = c(min(powerfactor_tidy$pfChart), max(powerfactor_tidy$pfChart)),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
  ggtitle("Mega Watt vs Power Factor (Shifted Value)" ) +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title = element_blank())

```

### Power Factor Shifted Value (Lagging and Leading)

```{r Power Factor Shifted Value By Hour, fig.width=10, fig.height=8}
ggplot(data = powerfactor_tidy, aes(x = h, y = pfChart, color=desc)) + 
  geom_point(size = 1, position = 'jitter') +
  scale_y_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  facet_wrap(~substationName) + 
  geom_hline(yintercept = 1, size = 0.2) +
  ggtitle('Power Factor Shifted Value (Lagging and Leading) By Substation') +
  xlab('hour') +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title = element_blank())

ggplot(data = powerfactor_tidy, aes(x = h, y = pfChart, color=desc)) + 
  geom_point(size = 0.5, position = 'jitter') +
  scale_y_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  facet_wrap(~dtReadDay) + 
  geom_hline(yintercept = 1, size = 0.2) +
  ggtitle('Power Factor Shifted Value (Lagging and Leading) By Day') +
  xlab('hour') +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title = element_blank())

```

### Power Factor Ratio Plot

```{r Power Factor Ratio Plot, fig.width=10, fig.height=6}
# Power Factor = kw / SQRT(kw^2 + kvar^2)  
ggplot(data=powerfactor_tidy[!is.na(powerfactor_tidy$mw), ], 
       aes(x=(mw), y=sqrt((mw*mw)+(mvar*mvar)))) +
  geom_point(position=position_jitter(h=1),aes(color=desc),alpha=1/5) + 
  ggtitle("Power Factor Ratio Plot\nMW and SQRT(kw^2 + kvar^2)" ) +
  theme_bw(base_family= 'Helvetica')  +
  theme(legend.title=element_blank()) 

```

### Voltage Scatter Plots With Bucket (Cut)

```{r Voltage Scatter Plot Bucket Group Facet, fig.width=12, fig.height=6}

ggplot(data = voltage_tidy, aes(x=h, y=VoltsHalf))+
  geom_point(aes(color=voltage.bucket), size=1, position='jitter', alpha=.75)+
  scale_y_continuous()+
  scale_color_manual(values=rev(brewer.pal(4,"Reds")),
                     guide = guide_legend(title="volts",
                                          reverse=T,
                                          overide.aes=list(size=3))) +
  ggtitle("Voltage Intervals Grouped") +
  xlab("hour")+
  ylab("voltage") +
  theme_bw(base_family= 'Helvetica') 
```

```{r Summary Of Voltage Bucket Table Data By Substation - substation_bucket_review}
# Review the total counts for voltage in each substation this goes with scatter plots 
# in bi or multivariate data

substation_bucket_review <-
  voltage_tidy %>% 
    group_by(substationName, voltage.bucket) %>% 
    summarise(count = n())  %>% 
        arrange(substationName, voltage.bucket)

# pivot the data for easy review
substation_bucket_review <- spread(substation_bucket_review, voltage.bucket, count)

```

```{r Voltage Scatter Plot by Bucket Facet per Substation, fig.width=12, fig.height=14}

ggplot(data = voltage_tidy, aes(x=h, y=VoltsHalf))+
  geom_point(aes(color=voltage.bucket), size=1, position='jitter')+
  facet_wrap(~substationName, ncol=3) +
  scale_y_continuous()+
  scale_color_manual(values=rev(brewer.pal(4,"Reds")),
                     guide = guide_legend(title="volts",
                                          reverse=T,
                                          overide.aes=list(size=3))) +
  ggtitle("Voltage Intervals by Bucket") +
  xlab("hour")+
  ylab("voltage") +
  theme_bw(base_family= 'Helvetica') 

ggplot(data = voltage_tidy, aes(x=h, y=VoltsHalf))+
  geom_point(aes(color=voltage.bucket), size=1, position='jitter')+
  facet_wrap(~dtReadDay, ncol=3) +
  scale_y_continuous()+
  scale_color_manual(values=rev(brewer.pal(4,"Reds")),
                     guide = guide_legend(title="volts",
                                          reverse=T,
                                          overide.aes=list(size=3))) +
  ggtitle("Voltage Intervals by Bucket") +
  xlab("hour")+
  ylab("voltage") +
  theme_bw(base_family= 'Helvetica') 


```

###### Voltage Bucket Summary 
Counts of the voltage instances in each bucketed range.
```{r Voltage Bucket Summary}
summary(voltage_tidy$voltage.bucket)
```

```{r Summary Of Voltage Bucket Table Data By Day - substation_bucket_review}
# Review the total counts for voltage in each substation this goes with scatter plots 
# in bivar. or multivariate data
substation_bucket_day_review <-
  voltage_tidy %>% 
    group_by(dtReadDay, substationName, voltage.bucket) %>% 
    summarise(count = n())  %>% 
        arrange(dtReadDay, substationName, voltage.bucket)

# pivot the data for easy review
substation_bucket_day_review <- spread(substation_bucket_day_review, voltage.bucket, count)

```

```{r substation_bucket_day_review add meta data to the dataset}
# calculate interval totals, TODO: need to check of .na values here
substation_bucket_day_review$total <- substation_bucket_day_review$`(0,114]` +
  substation_bucket_day_review$`(114,120]` +
  substation_bucket_day_review$`(120,126]` +
  substation_bucket_day_review$`(126,140]`

# calculate the percent for each bucket
substation_bucket_day_review$p1 <- 
  100*(substation_bucket_day_review$`(0,114]`/substation_bucket_day_review$total)

substation_bucket_day_review$p2 <- 
  100*(substation_bucket_day_review$`(114,120]`/substation_bucket_day_review$total)

substation_bucket_day_review$p3 <- 
  100*(substation_bucket_day_review$`(120,126]`/substation_bucket_day_review$total)

substation_bucket_day_review$p4 <- 
  100*(substation_bucket_day_review$`(126,140]`/substation_bucket_day_review$tota)

```

##### Voltage Bucket Data By Day
```{r Summary Of Voltage Bucket Table Data By Day - substation_bucket_review table}
# output to table
kable(substation_bucket_day_review)

```


#### Compare Voltage and Power Factor Range Counts
```{r Totalize up the pf ranges}
# MIGHT NEED THIS BY SUBSTATION AS WELL 
powerfactor_ranges_day <-
  powerfactor_tidy %>% 
    group_by(dtReadDay, substationName, pf.range) %>% 
    summarise(count = n())  %>% 
        arrange(dtReadDay, substationName, pf.range)

# pivot the data for easy review
powerfactor_ranges_day <- spread(subset(powerfactor_ranges_day, !is.na(pf.range)), pf.range, count)

```

```{r DEAD END}
range_counts <- left_join(powerfactor_ranges_day,
                          substation_bucket_day_review, 
                          by= c("dtReadDay", "substationName"))

# subset the columns I want
range_counts <- range_counts[,1:13]

test <- gather(range_counts, "pf.range", "pf.n", 3:9)
test <- gather(test, "v.range", "v.n", 3:6)

ggplot(data = subset(test,!is.na(v.n)), aes(x=v.range, y=v.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) 

ggplot(data = subset(test,!is.na(v.n)), aes(x=v.range, y=v.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) +
  facet_wrap(~substationName)

ggplot(data = subset(test,!is.na(v.n)), aes(x=v.range, y=v.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) +
  facet_wrap(~dtReadDay)

###

ggplot(data = subset(test,!is.na(pf.n)), aes(x=pf.range, y=pf.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) 

ggplot(data = subset(test,!is.na(pf.n)), aes(x=pf.range, y=pf.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) +
  facet_wrap(~substationName)

ggplot(data = subset(test,!is.na(pf.n)), aes(x=pf.range, y=pf.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) +
  facet_wrap(~dtReadDay)

###
ggplot(data = subset(test,!is.na(pf.n)), aes(x=pf.range, y=pf.n))+
  geom_point(position=position_jitter(w=0.3, h=1), alpha=1/5) 


range_totals <-
  test %>% 
    group_by(dtReadDay, substationName, pf.range, v.range) %>% 
    summarise(v.n = sum(v.n, na.rm=TRUE),
              pf.n= sum(pf.n,na.rm=TRUE))  %>% 
        arrange(dtReadDay, substationName, pf.range,v.range)



```

```{r LAST HOP}
#ggplot(data = substation_bucket_day_review, aes(x = dtReadDay, y=p1)
```

This chart demonstrates how the voltage can vary on the lower and upper ends of the voltage ranges.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest?

* MegaWatt vs Power Factor (GOOD)  
* Voltage vs Power Factor  (NOTHING)  
* Voltage vs PF with Linear Model - Facet Substation  (NOTHING)  
* MegaWatt vs PF with Linear Model by Substation (GOOD)  
* Power Factor Shifted Value (Lagging and Leading) By Hour (GOOD), FINAL CHART  
* Power Factor Ratio Plot (GOOD)  

### Were there any interesting or surprising interactions between features?

#### Voltage Scatter Plots With Bucket and Power Factor 
The voltage and power factor for the entire system seem to track or 
follow a similar trend here.  They have similar shapes.  Or at least
when the power factor range increases the voltage drops less in
the system.  This could be for many reasons.

```{r Compare Voltage and Power Factor}
voltagehour <-
  ggplot(data = voltage_tidy, aes(x=h, y=VoltsHalf))+
  geom_point(aes(color=voltage.bucket), size=1, position='jitter', alpha=.75)+
  scale_y_continuous()+
  scale_color_manual(values=rev(brewer.pal(4,"Reds")),
                     guide = guide_legend(title="volts",
                                          reverse=T,
                                          overide.aes=list(size=3))) +
  xlab("")+
  ylab("voltage") +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.position="none") +
 theme(axis.text.x=element_blank(),
   axis.ticks.x=element_blank(),
   #plot.margin = rep(unit(0,"null"),4),
   #panel.margin = unit(0,"null"),
   axis.ticks.length = unit(0,"null"),
   axis.ticks.margin = unit(0,"null")) +
 labs(x=NULL)

pfhour <- 
  ggplot(data=powerfactor_tidy, aes(x=h, y=pfChart)) +
  geom_point(size=1) + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  geom_hline(yintercept=1) +
  ylab("power factor") +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.position="none")

p1 <- voltagehour + scale_x_continuous(limits=c(0, 23))
p2 <- pfhour + scale_x_continuous(limits=c(0, 23))

grid.arrange(p1, p2, ncol=1)
```


# Models
One idea to be explored is if we can predict when the power factor will 
go below 0.98 lagging or leading.  This could help us change the system parameters to keep the network in operating efficiency.
## Goals
* Create model for each substation to predict when power factor goes below 0.98 based on the
hourly readings.  If we can predict at a minimum 2 hours ahead this would be idea. The initial

```{r Linear Model using powerfactor_sort }
# pf = kw / SQRT(kw^2 + kvar^2)  
# m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = powerfactor_columns_sort)
# how about adding voltage as well... 
m1 <- lm(pf ~ I(sqrt(mw)), data = powerfactor_sort)
m2 <- update(m1, ~ . + mw)
m3 <- update(m1, ~ . + mvar)
m4 <- update(m2, ~ . + h)
m5 <- update(m3, ~ . + desc)

mtable(m1, m2, m3, m4, m5)
```


## Linear Model Test
Expected Result : 0.9903636
```{r Linear Model Test}
load = data.frame(mw = 16.140,
                  mvar = 2.257,
                  h = 22, 
                  desc="Lagging")


pf_estimate = predict(m5, newdata = load,
                        interval="prediction", level = .95)

pf_estimate
```

### Summary
```{r lm summary}
summary(m5)
```

### Anova
```{r lm anova}
anova(m5)
```

### Residual Plots
```{r Residuals Plot}
plot(m5)
```



## Discuss the strengths and limitations of your model.

# Final Plots and Summary
Are the final three plots varied and do they meet some of the following criteria:  
1 Draw comparisons.    
2 Identify trends.    
3 Engage a wide audience.  
4 Explain a complicated finding.  
5 Clarify a gap between perception and reality.  
6 Enable the reader to digest large amounts of information.  


Each plot reveals an important and different comparison or trend in the data. The plots incorporate many of the variables from the data set in a way that allows the plots to convey a lot of information while still being interpreted easily. The plots fulfill 4 or more of the criteria.

### Plot One
Voltage and Power Factor Outlier plot helps the user to quickly identify 
which substations are out the bounded region. 

```{r Plot_One - Voltage and Power Factor Outlier Plot, fig.width=12, fig.height=9}
# maybe facet both by day?
# Main Chart
voltage_pf_plot <- 
  ggplot(data=summary_total_day[!is.na(summary_total_day$pf.median), ], 
       aes(x=pf.median, y=v.median)) +
    geom_point(size=3, color='green') +
    facet_wrap(~dtReadDay)

# Add lagging outliers if they exist
if (!(is.data.frame(lagging_outliers) && nrow(lagging_outliers)==0)) {
voltage_pf_plot <-
  voltage_pf_plot +
    geom_point(data=lagging_outliers,
               aes(x=pf.median, y=v.median), color='red', size=4) +
    geom_text(data=lagging_outliers,
              aes(label=substationName),hjust=0, vjust=2,size=3) +
    geom_errorbar(data=lagging_outliers,
                  aes(ymin = v.min, ymax = v.max), width = 0.0025, alpha=1/2) +
    geom_errorbarh(data=lagging_outliers,
                   aes(xmin = pf.min,xmax = pfc.max, height = .4), alpha=1/2)
}
# Add leading outliers if they exist
if (!(is.data.frame(leading_outliers) && nrow(leading_outliers)==0)) {
  voltage_pf_plot <-
    voltage_pf_plot +
      geom_point(data=leading_outliers,
                 aes(x=pf.median, y=v.median), color='red', size=4) +
      geom_text(data=leading_outliers,
                aes(label=substationName),hjust=0, vjust=2,size=3) +
      geom_errorbar(data=leading_outliers,
                    aes(ymin = v.min, ymax = v.max), width = 0.0025, alpha=1/2) +
      geom_errorbarh(data=leading_outliers,
                   aes(xmin = pf.min,xmax = pfc.max, height = .4), alpha=1/2)
}

# Additional Formating
voltage_pf_plot <-
  voltage_pf_plot +
      scale_y_continuous(limits = c(110, 130)) +
      scale_x_continuous(limits = c(0.94, 1.06),
                         breaks=c(.94,.96,.98,1,1.02,1.04,1.06),
                         labels=c(".94", ".96", ".98","1", ".98", ".96",".94"))
voltage_pf_plot <-
  voltage_pf_plot +
      geom_segment(aes(x = 0.98, y = voltageLow, xend = 0.98, 
                       yend = voltageHigh), size=0.15,colour='black') + # left
      geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                       yend = voltageHigh), size=0.15,colour='black') + # right
      geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                       yend = voltageHigh), size=0.05,colour='black') + # top
      geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                       yend = voltageLow), size=0.15,colour='black')   # bottom

    
voltage_pf_plot +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica')  +
    ggtitle("Voltage & Power Factor Summary" )




```

```{r Plot_One - Voltage and Power Factor By Hour By Day, fig.width=12, fig.height=9}
ggplot(data=summary_total[!is.na(summary_total$pfChart), ], 
       aes(x=pfChart, y=v.median)) +
    geom_point(aes(color=substationName),
               position=position_jitter(h=1/2),
               alpha=1/1, size=2) + 
    facet_wrap(~dtReadDay) +
    scale_fill_brewer(palette="Set2") +  
    scale_y_continuous(limits = c(110, 130)) +
      scale_x_continuous(limits = c(0.94, 1.06),
                         breaks=c(.94,.96,.98,1,1.02,1.04,1.06),
                         labels=c(".94", ".96", ".98","1", ".98", ".96",".94")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black') + # bottom
    ggtitle("Voltage & Power Factor By Hourly" ) +
    ylab("median voltage") +
    xlab("power factor") +  
    theme_bw(base_family= 'Helvetica')  +
    theme(legend.title = element_blank())

```


```{r Voltage Power Factor All Substations,fig.width=12, fig.height=6}
# Remove all the outliers from the main dataset
# This plot is an example of trying to view all datapoints in the
# required customers scope.  Going to try and split this into
# groups
ts <- summary_total_day
subsOne <-  ts[(ts$substationName %in% substations[1:6,2]),]
subsTwo <-  ts[(ts$substationName %in% substations[7:13,2]),]
subsThree <-  ts[(ts$substationName %in% substations[14:20,2]),]
subsFour <- ts[(ts$substationName %in% substations[21:24,2]),]

# TODO - It would be nice to reduce some code here
plot1 <- ggplot(data=subsOne[!is.na(subsOne$pfc.median), ], 
       aes(x=pfc.median, y=v.median,color=substationName)) +
    geom_point(size=3) + 
    geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    geom_errorbar(aes(ymin = v.min, ymax = v.max), width = 0.0025,alpha=1/2) + 
    geom_errorbarh(aes(xmin = pfc.min, xmax = pfc.max,  height = .4),alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black')  +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank())

plot2 <- ggplot(data=subsTwo[!is.na(subsTwo$pfc.median), ], 
       aes(x=pfc.median, y=v.median,color=substationName)) +
    geom_point(size=3) + 
    geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    geom_errorbar(aes(ymin = v.min, ymax = v.max), width = 0.0025,alpha=1/2) + 
    geom_errorbarh(aes(xmin = pfc.min, xmax = pfc.max,  height = .4),alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black')  +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom")+
    theme(legend.title=element_blank())

plot3 <- ggplot(data=subsThree[!is.na(subsThree$pfc.median), ], 
       aes(x=pfc.median, y=v.median,color=substationName)) +
    geom_point(size=3) + 
    geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    geom_errorbar(aes(ymin = v.min, ymax = v.max), width = 0.0025,alpha=1/2) + 
    geom_errorbarh(aes(xmin = pfc.min, xmax = pfc.max,  height = .4),alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black')  +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank())

plot4 <- ggplot(data=subsFour[!is.na(subsFour$pfc.median), ], 
       aes(x=pfc.median, y=v.median,color=substationName)) +
    geom_point(size=3) + 
    geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    geom_errorbar(aes(ymin = v.min, ymax = v.max), width = 0.0025,alpha=1/2) + 
    geom_errorbarh(aes(xmin = pfc.min, xmax = pfc.max,  height = .4),alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + 
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black')  +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank())

grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)

```

```{r summary_total_day table data, include=FALSE}

kable(summary_total_day[c("dtReadDay",
                          "substationName",
                          "v.min",
                          "v.median",
                          "v.min",
                          "v.max",
                          "pfc.min",
                          "pfc.median",
                          "pfc.max",
                          "desc")],digits=2)

```

This chart represents the voltage and power factor for all data points 
collected for `r report_date`. The chart allows one to see how much 
of the data falls out of the voltage and power factor defined region 
(`r pfLow` (left) to `r pfHigh` (right), and `r voltageLow`v to `r voltageHigh`v).

We can see we have more points with Lagging power factors outside the 
required range for efficient power factor values.

NEED 4 or more... 

1 Draw comparisons. 
 * YES - Comparisons between substations
2 Identify trends.    
 * NO - Trends - Nope, need multiple days... or day without VVar day with VVar
3 Engage a wide audience.
 * Yes, easy to review...
4 Explain a complicated finding.  
 * YES - which substations behaved and what where their min and max for pf and voltages
5 Clarify a gap between perception and reality.  
 * MAYBE - If a person looks at the current momentary... status
 * the daily status might be different
6 Enable the reader to digest large amounts of information. 
  * YES - Power Factor, Voltage, Ranges, inside the comfort zone, 


```{r Plot_One Voltage and Power Factor Scatter Plot, include = FALSE, fig.width=12, fig.height=6}
# using total b/c it is both voltage and powerfactor
ggplot(data=summary_total[!is.na(summary_total$pfChart), ], 
       aes(x=pfChart, y=v.median)) +
    geom_point(aes(color=substationName),
               position=position_jitter(h=1/2),
               alpha=1/1, size=2) + 
    scale_fill_brewer(palette="Set2") +  
    scale_y_continuous(limits = c(100, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(.94,.96,.98,1,1.02,1.04,1.06),
                       labels=c(".94", ".96", ".98","1", ".98", ".96",".94")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black') + # bottom
    ggtitle("Substation Voltage & Power Factor By Hour - Medians" ) +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica')  +
    theme(legend.title = element_blank())

```


### Plot Two
Heatmap of Voltage to help user to see how all substations performed 
during each 15 minute period during the day.  One chart helps to
see who is in the red (low) or in the purple (high) voltage ranges.

```{r Plot_Two Voltage Heatmaps 15min, fig.width=12, fig.height=8}
# Note we can have multiple meters per 15min intervals which means multiple points
# stacked on each other
heatmap_voltage_15min <-
   ggplot(voltage_tidy, aes(y=substationName, x=hm, fill=VoltsHalf))

heatmap_voltage_15min +
 geom_tile(colour="white") +
 scale_fill_gradient2(
                      midpoint=122,
                      limits=c(100
                               ,130)) +
 theme_bw(base_family= 'Helvetica')  +
 theme(axis.text.x = element_text(angle = -90, hjust = 0, size = 8))  +
 theme(legend.title=element_blank()) +
 xlab("") +
 ylab("") +
 ggtitle("Voltage 15min for each hour \n(multiple meters)" )
```

```{r Plot_Two Voltage Heatmaps Hour, fig.width=12, fig.height=6, include=FALSE}
# Keeping and turning this plot off
 heatmap_voltage_hour_min <-
   ggplot(summary_voltage, aes(y=substationName, x=h, fill=v.min))

heatmap_voltage_hour_min + 
 #geom_raster() +  
 geom_tile(colour="white") +
 scale_fill_gradient2(
                      midpoint=mean(summary_voltage$v.min),
                      limits=c(min(summary_voltage$v.min)
                               ,max(summary_voltage$v.min))) +
 theme_bw(base_family= 'Helvetica')  +
 theme(axis.text.x = element_text(angle = -90, hjust = 0))  +
 theme(legend.title=element_blank()) +
 xlab("") +
 ylab("") +
 ggtitle("Voltage minimum for each hour" )

```

```{r Plot_Two Power Factor Heatmaps,fig.width=12, fig.height=6, include=FALSE}
library(scales)
heatmap_powerfactor_hour <-ggplot(powerfactor_tidy, aes(y=substationName, x=h, fill=pfChart))

heatmap_powerfactor_hour +
  geom_tile() +
   #scale_x_continuous(breaks = seq(0,23,step= 1)) +  
   #scale_x_discrete(breaks=c("04:00","08:00","12:00","16:00","20:00","23:00")) +
   # scale_fill_gradient2(
   #                      midpoint=mean(powerfactor_tidy$pfChart),
   #                      limits=c(min(powerfactor_tidy$pfChart)
   #                               ,max(powerfactor_tidy$pfChart))) +
   scale_colour_gradient2(low="red", high="blue") +
   theme(legend.title=element_blank()) +
   xlab("") +
   ylab("")

```

1 Draw comparisons.    
2 Identify trends.    
3 Engage a wide audience.  
4 Explain a complicated finding.  
5 Clarify a gap between perception and reality.  
6 Enable the reader to digest large amounts of information. 


### Plot Three

Description goes here please... Note, I think I have a better plot to put here, this
one kinda goes along with Plot_One, it is just another form of the plot.

```{r Plot Three, fig.width=12, fig.height=8}
ggplot(data = powerfactor_tidy, aes(x = h, y = pfChart, color=desc)) + 
  geom_point(size = 0.5, position = 'jitter') +
  scale_y_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +  
  facet_wrap(~substationName) + 
  geom_hline(yintercept = 1, size = 0.2) +
  ggtitle('Power Factor Shifted Value (Lagging and Leading)') +
  xlab('hour') +
  theme_bw(base_family= 'Helvetica') +
  theme(legend.title = element_blank())
```



Total

# Reflection
The section explains any important decisions in the analysis and how those 
decisions affected the analysis.

Need to make sure to handle cases when data is missing from a day 
over a few date ranges.

Automate additional processes.

## Voltage
Voltage seems to be normally like distributed by the hour through out the 
day and by substation for the entire by hour.  This seems good since
large flucuations in voltage are bad for consumers and commercial 
businesses.

Power Factor can flucuation from lagging to leading in on substation in
a single day.  This requires more effort to control the voltage
, watts, and vars across the power lines.  Can I show
that when voltage is tightly controled we have less variability in 
the power factor??????

TODO: Find a sub with the best range in voltage and look at its power
factor.  Then compare it's MW and MVARS to all other substations.

The section reflects on how the analysis was conducted and reports on the 
struggles and successes throughout the analysis. The section provides at 
least one idea or question for future work.

The section provides a rich and well-written reflection of 
  * struggles
    * Date Formatting
    * Exploring for relationships

  * successes
    * MegaWatt vs MegaVar
    * MegaWatt vs Power Factor
    * Drilling down from Substation to Meter to find the bad meters

  * Lessons learned
    Electricity is a pain to review.

The section poses ideas or questions for future work. 
  * multiple days would be benficial 
  * months to months
  * 12 month analysis of trends
  * seasonal trends
  * collect more power factor data down to 15min intervals


# Equations
## Ohm's Law
We monitor voltage since a utility can change the voltage across the 
power lines.  Keeping the system balanced at low voltages between 114 
and 126 helps to improve the efficeny of the power transmission since
lower voltages mean less resistance.

> I = Current  
  R = Resistance  
  V = Voltage  
  
$V = I*R$  

$R = \frac{V}{I}$  

$I = \frac{V}{R}$

## Calculate watts from Voltage

> P(kW) = PF × I(A) × V(V) / 1000

Power - kW, kiloWatts  
Power Factor  
I - Amps  
Voltage - Volts  

3 Phase  
P(kW) = √3 × PF × I(A) × VL-L(V) / 1000  

## Power Factor 
P, Real Power - kW -> PMWD3D (MW)
Real power is kilowatts, in the initial dataset this is represented as PMWD3D.

S, Apparent Power    
We will be solving for apparent power.

Q, Reactive Power 
Reactive power is kVAR, in the initial dataset PMQD3D is Delivered and  
PMQR3D is Received.  The total kVars are (PMQD3D-PMQR3D).  

### The Power Triangle Equation
[See Reference for explanation](https://en.wikipedia.org/wiki/Power_factor)  

$S^2 = P^2 + Q^2$  

$S=\sqrt{P^2 + Q^2}$

### Power Factor
The power factor is defined as the ratio of real power to apparent power.  
 
$$Power Factor=\frac{P}{\sqrt{P^2 + Q^2}}$$

$$Power Factor =\frac{kW}{\sqrt{kW^2 + (kVAR Delieverd - kVAR Received)^2}}$$

### Leading and Lagging Power Factor
* Delivered VARs - Received VARs   > 0 Lagging  
* Delivered VARs - Received VARs   < 0 Leading

# References
https://en.wikipedia.org/wiki/Power_factor  
https://en.wikipedia.org/wiki/Ohm%27s_law  
https://en.wikipedia.org/wiki/Volt-ampere_reactive  
http://www.statpower.net/Content/310/R%20Stuff/SampleMarkdown.html  
http://rmarkdown.rstudio.com/authoring_basics.html  
http://www.ats.ucla.edu/stat/r/library/contrast_coding.htm   
http://vita.had.co.nz/papers/tidy-data.pdf 
http://adv-r.had.co.nz/Style.html

https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/
http://www.statmethods.net/management/aggregate.html
http://yihui.name/knitr/options/
http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html
http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html
http://data.princeton.edu/R/linearModels.html 
https://statistics.laerd.com/statistical-guides/pearson-correlation-coefficient-statistical-guide.php

